---
title: "workflow_viability_selection"
output:
  pdf_document: default
  html_document:
    toc: yes
---
# Library
```{r}
#install.packages("DiagrammeR")
library(DiagrammeR)
```
# Overall Workflow
```{r}
mermaid("
  graph TD
  A[Master Data Sheet] --calculate--> B[Metabolic Rates CO2 production per hr per mass]
  B --> T[Day 11]
  B --> R[Day 15]
  A -- calculate --> C[lifespan]
  T --> E[Scatter plot showing relationship between life span and MR]
  R --> E
  C --> E
        ")

```

##Detailed workflow for calculating metabolic rate
How do we calculate metabolic rate?
Metabolic Rate = CO2 production/ hours
Mass Specific Metabolic Rate = CO2 production/hrs per mass

What are cohorts?
* Essentially handling replicats because we cannot sample all flies at once. For each cohort, there are two batches: (1) licor 7000 and (2) licor 6262
* To process more samples, measurements were spread across 2 respirometers(licors).
* Each licor has associated tape color indicated in tape column
* When purged (allowing pupae to breathe in a syringe under standardized air without CO2), there was a start and end time for beginning and end of purge for each tape samples (licor batch). 

What columns do we need from the master data sheet?
* mass_day10: mass on day 10 (mgs)
* purge_time_1: initial time for pupae to breathe in tube
* resp_time_1: time of respirometry sampling
* resp_day_11: respirometry on day 11

**Problem** with purge_time_1: only start and stop times except for apple cohort 1 
**Solution** create time sequence for each cohort from start to end


```{r}
mermaid("
  graph TD
  A[Master Data Sheet] --> B[Create time sequence for purge_time_1 in hrs minutes]
  B--subtract from 24 and add to resp_time_1 --> C[Total hrs]
  C --divide over resp_day11--> E[metabolic rate]
  C --multiply by mass and divide over resp_day11--> D[Mass specific metabolic rate]
")
```

## Workflow for Merging Pupal Death Data with Lifespan Data
```{r}
mermaid("
  graph TD
  A[Master Data Sheet] --filter out the non-eclosers--> B[Subset of ind. who haven't eclosed]
  A -- adult death date - eclosion date --> E[Lifespan Data]
  C[Pupal Deaths Data Sheet]
  B --merge based on host, cohort_day, well_id, uniqueID --> D[Merged Data Sheets]
  C --merge based on host, cohort_day, well_id, uniqueID --> D
  D --merge--> F[Final Data Sheet]
  E --merge--> F[Final Data Sheet]
  F --analyses using R --> G[negative binomial regression]
  F --analyses using R package survival --> H[Survival analysis]
  F --analyses using R--> I[COX regression]
")
```

# Session Info
```{r}
sessionInfo()
```