---
title: "viability_selection_mr_lifespan"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
---

#Libraries
```{r}
library(dplyr) #data parsing library
library(data.table) #data parsing library mainly for reading and writing out
library(ggplot2) #data visualization library
library(lubridate) #package to deal with time variables
library(curl) #incorporate URLs
library(DiagrammeR) #figure aesthetics
library(MASS) #Functions and datasets to support
library(BBmisc) #Miscellaneous functions
library(survival) #survival data analysis
library(survminer) #survival data analysis
library(gridExtra) #figure aesthetics
```

# Download data and do preliminary parsing
Data are publicly available via [github](https://raw.githubusercontent.com/adnguyen/Circadian_rhythm_runs_seasonal_timing/master/Data/2018-05-30_rhagoletis_masterdata_data_slice.csv)
```{r}
data<- fread("https://raw.githubusercontent.com/adnguyen/Circadian_rhythm_runs_seasonal_timing/master/Data/2018-05-30_rhagoletis_masterdata_data_slice.csv")

#look at summary of data
glimpse(data)
```

#Calculating metabolic rates
Metabolic Rate = CO2 production/ hours
Mass Specific Metabolic Rate = CO2 production/hrs per mass
**Problem** with purge_time_1: only start and stop times except for apple cohort 1
**Solution** create time sequence for each cohort from start to end and merge into master data sheet
##Day 11
###Creating the time sequence
```{r}
glimpse(data$purge_time_1)
hm(data$purge_time_1)
data$day10purge <- lubridate::hour(hm(data$purge_time_1))+lubridate::minute(hm(data$purge_time_1))/60

#Obtaining start and end (min and max) of purges and smaple size for each host, cohort day, and tape
param <- data%>%
  group_by(cohort_day, tape)%>%
  summarise(max=max(day10purge, na.rm=TRUE), min=min(day10purge, na.rm=TRUE), n=length(cohort_day))

#goal: for this section, we want a sequence of times for day 10 purge
data2 <- data%>%
  group_by(cohort_day, tape)%>%
  mutate(.,day10purge.trans=seq(from = min(day10purge, na.rm=TRUE), to = max(day10purge, na.rm=TRUE), length.out = length(Host)))
glimpse(data2)

#cohort 2&3 for apple have the right sequence of purge times so they(day10purge) need to be replaced into corresponding day10purge.trans column
  #extract number of rows we want to replace
data2[1:500,]%>%
  filter(cohort_day < 4)%>%
  dim()
data2[1:211,46]<- data2[1:211,45]
data2[1:211,46]
```

###Calculating start and end time for total amount of hours of CO2 production
```{r}
glimpse(data$resp_time_1)
hms(data$resp_time_1)
data2$day10resp <- lubridate::hour(hms(data$resp_time_1))+lubridate::minute(hms(data$resp_time_1))/60

#reformatting to obtain positive values
data2$total_time_day10 <- (24 - data2$day10purge.trans) + data2$day10resp
```

###Metabolic Rate Calculation
```{r}
#getting denominator for mass-specific mr
data2$total_time_day10 * data$mass_day10

#Metabolic Rate for day 11
data2$MR11<- data$resp_day11/(data2$total_time_day10)

#Mass specific Metabolic Rate for day 11
data2$msMR11<- data$resp_day11/(data2$total_time_day10 * data$mass_day10)
```

#Controlling for blank controls
```{r}
#Assign the blanks to an object
data3 <- data2%>%
  group_by(cohort_day, tape)%>%
  filter(Site_name=="Blank")%>%
  summarise(mean.blank=mean(MR11,na.rm=TRUE))

#check columns
glimpse(data3)
data3$mean.blank

#merge data3 and data2 by cohort day and tape and create a mean blanks column
data4 <- inner_join(data2, data3, by=c("cohort_day", "tape"))
data4$mean.blank

#do some corrections
data5 <- data4%>%
  mutate(MR11.cor = MR11 - mean.blank, msMR11.cor = msMR11 - mean.blank)
glimpse(data5)

data5.neg <- data5%>% 
  filter(MR11.cor<0)%>%
  dim()
```

## Figure of Metabolic Rate between Hosts
```{r}
#Boxplot
ggplot(data5, aes(x=Host, y=MR11.cor))+geom_boxplot()

#Histogram
ggplot(data5, aes(x=MR11.cor, fill=(Host)))+geom_histogram(position = "identity", alpha=.5)
```

###Calculate eclosion days
```{r}
data5$neweclosions<-difftime(as.Date(data5$eclosion_date), as.Date(data5$Eclosion_reference_date),units=c("days"))+15
```

###Figures of eclosion and metabolic rate data
```{r}
data5.treatsub<-data5%>%
  filter(treatment!="GC"&treatment!="")

#Histogram comparison between MR of RT & SO
ggplot(data5.treatsub, aes(x=MR11.cor, fill=Host))+geom_histogram(position = "identity", alpha=.5)+facet_grid(.~treatment)

#Density comparison between MR of RT & SO
ggplot(data5.treatsub, aes(x=MR11.cor, fill=Host))+geom_density(position = "identity", alpha=.5)+facet_grid(.~treatment)

#Histogram comparison between eclosion date of RT & SO
ggplot(data5.treatsub, aes(x=neweclosions, fill=Host))+geom_histogram(position = "identity", alpha=.5)+facet_grid(.~treatment)

#Density comparison between eclosion date of RT & SO
ggplot(data5.treatsub, aes(x=neweclosions, fill=Host))+geom_density(position = "identity", alpha=.5)+facet_grid(.~treatment)
```

###Test differences in eclosion timing between host and experiment
Using ANOVA which tests differences between two or more means
* The null hypothesis would be that the two means are equal
* Significance would indicate two means are not equal
*_One Way ANOVA_* compares two means from two independent groups, in this case eclosion date and host (specific to SO and RT)

*_Two Way ANOVA_* compares means of two independent variables affecting one dependent variable, in this case the effect of Host and treatment (interacting) on eclosion
```{r}
#Focus only on RT and SO treatments
data5.treatsub<-data5%>%
  filter(treatment!="GC"&treatment!="")

#Two Way ANOVA
mod1<- aov(neweclosions ~ Host*treatment, data=data5.treatsub)
summary(mod1)

#narrow down to SO & RT
data5.SO<-data5.treatsub%>%
  filter(treatment=="SO")

data5.RT<-data5.treatsub%>%
  filter(treatment=="RT")

#One Way ANOVA for RT & SO
mod2RT<-aov(neweclosions~Host, data=data5.RT)
summary(mod2RT)

mod2SO<-aov(neweclosions~Host, data=data5.SO)
summary(mod2SO)

#how balanced it is
data5.RT%>%
  group_by(Host)%>%
  summarise(n=length(Host))

```

###Explore relationship between eclosion days and MR
**scatter plots**
```{r}
ggplot(data5.treatsub,aes(x=MR11.cor, y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="loess")+facet_grid(.~treatment)

#plotting mr11 on the x axis and new eclosions on the y axis; color points by host, fit curve to data
ggplot(data5.treatsub,aes(x=MR11.cor, y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="loess")

#same - fit linear curve to data
ggplot(data5.treatsub,aes(x=MR11.cor, y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="lm")

#could log transform mr to make it more linear
ggplot(data5.treatsub,aes(x=log10(MR11.cor), y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="lm") + xlim(-3.2,-.5)
```

###Calculating lifespan
```{r}
data5.treatsub$lifespan<-difftime(as.Date(data5.treatsub$Adult_death_date, na.rm=TRUE), as.Date(data5.treatsub$eclosion_date, na.rm=TRUE),units=c("days"))

```

###Figures of lifespan between hosts
```{r}
ggplot(data5.treatsub, aes(x=Host, y=lifespan))+geom_boxplot()
ggplot(data5.treatsub, aes(x=lifespan, fill=(Host)))+geom_histogram(position = "identity", alpha=.5)
```

###Figures Associating MR with lifespan
```{r}
#Loess
ggplot(data5.treatsub,aes(x= MR11.cor, y=lifespan, colour=Host))+geom_point()+stat_smooth(method="loess") + facet_grid(.~treatment)

#Linear
ggplot(data5.treatsub,aes(x=MR11.cor, y=lifespan, colour=Host))+geom_point()+stat_smooth(method="lm") + facet_grid(.~treatment)

#Taking the log
ggplot(data5.treatsub,aes(x=log10(MR11.cor), y=lifespan, colour=Host))+geom_point()+stat_smooth(method="loess") + xlim(-3.2,-.5) + ylim(-.5, 150)
```

## Repeat above, but with day 15, starting with time sequence
```{r}
glimpse(data$purge_time_2)
hm(data$purge_time_2)
data$day15purge <- lubridate::hour(hm(data$purge_time_2))+lubridate::minute(hm(data$purge_time_2))/60

#Getting start and end (min and max) of purges and sample size for each host, cohort day, and tape
param <- data%>%
  group_by(cohort_day, tape)%>%
  summarise(max=max(day15purge, na.rm=TRUE), min=min(day10purge, na.rm=TRUE), n=length(cohort_day))

#goal: for this section, we want a sequence of times for day 15 purge
data2.15 <- data%>%
  group_by(cohort_day, tape)%>%
  mutate(.,day15purge.trans=seq(from = min(day15purge, na.rm=TRUE), to = max(day15purge, na.rm=TRUE), length.out = length(Host)))
glimpse(data2.15)
```

###Calculating start and end time for total amount of hours of CO2 production
```{r}
glimpse(data$resp_time_2)
hms(data$resp_time_2)
data2.15$day15resp <- lubridate::hour(hms(data$resp_time_2))+lubridate::minute(hms(data$resp_time_2))/60

#Obtaining total time in hours
data2.15$total_time_day15 <- (24 - data2.15$day15purge.trans) + data2.15$day15resp
```

###Metabolic Rate Calculation
```{r}
#getting denominator for mass-specific mr
data2.15$total_time_day15 * data$mass_day14

#Metabolic Rate
data2.15$MR15<- data$resp_day15/(data2.15$total_time_day15)

#Mass specfic metabolic rate
data2.15$msMR15<- data$resp_day15/(data2.15$total_time_day15 * data$mass_day14)
```

###Controlling for Blanks
```{r}
data3.15 <- data2.15%>%
  group_by(cohort_day, tape)%>%
  filter(Site_name=="Blank")%>%
  summarise(mean.blank2=mean(MR15,na.rm=TRUE))

#check columns
glimpse(data3.15)
data3.15$mean.blank2

#merge data3 and data2 by cohort day and tape
data4.15 <- inner_join(data2.15, data3.15, by=c("cohort_day", "tape"))
data4.15
data4.15$mean.blank2

#do some corrections
data5.15 <- data4.15%>%
  mutate(MR15.cor = MR15 - mean.blank2, msMR15.cor = msMR15 - mean.blank2)
glimpse(data5.15)

#Filter out negatives
data5.15.neg <- data5.15%>% 
  filter(MR15.cor<0)
```

###Calculate Eclosions Dates
```{r}
#Focus only on RT and SO treatments
data5.15.treatsub<-data5%>%
  filter(treatment!="GC"&treatment!="")

data5.15$neweclosions <- difftime(as.Date(data5.15$eclosion_date), as.Date(data5.15$Eclosion_reference_date),units=c("days"))+15

data5.15.treatsub$neweclosions<-difftime(as.Date(data5.15.treatsub$eclosion_date), as.Date(data5.15.treatsub$Eclosion_reference_date),units=c("days"))+15
```

###Test differences in eclosion timing between host and experiment for Day 15
```{r}
#Two Way ANOVA
mod1<- aov(neweclosions ~ Host*treatment, data=data5.15.treatsub)
summary(mod1)

#narrow down to SO & RT
data5.15SO<-data5.15.treatsub%>%
  filter(treatment=="SO")

data5.15RT<-data5.15.treatsub%>%
  filter(treatment=="RT")

#One Way ANOVA for RT & SO
mod2RT15<-aov(neweclosions~Host, data=data5.15)
summary(mod2RT15)

mod2SO15<-aov(neweclosions~Host, data=data5.15)
summary(mod2SO15)

#how balanced it is
data5.SO%>%
  group_by(Host)%>%
  summarise(n=length(Host))
```

### Figures looking at MR between hosts
```{r}
ggplot(data5.15, aes(x=Host, y=MR15.cor))+geom_boxplot()
ggplot(data5.15, aes(x=MR15.cor, fill=(Host)))+geom_histogram(position = "identity", alpha=.5)
```

###Explore relationship between eclosion days and MR
**scatter plots**
```{r}
data5.15.treatsub<-data5.15%>%
  filter(treatment!="GC"&treatment!="")

ggplot(data5.15.treatsub, aes(x=MR15.cor, y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="loess")+facet_grid(.~treatment)

#plotting mr15 on the x axis and new eclosions on the y axis; color points by host, fit curve to data
ggplot(data5.15.treatsub,aes(x=MR15.cor, y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="loess")

#same - fit linear curve to data
ggplot(data5.15.treatsub,aes(x=MR15.cor, y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="lm")

#could log transform mr to make it more linear
ggplot(data5.15.treatsub,aes(x=log10(MR15.cor), y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="lm") + xlim(-3.2,-.5)
```

###Figures associating MR with lifespan
```{r}
#mr15 w lifespan
data5.15.treatsub$lifespan<-difftime(as.Date(data5.15.treatsub$Adult_death_date, na.rm=TRUE), as.Date(data5.15.treatsub$eclosion_date, na.rm=TRUE),units=c("days"))

ggplot(data5.15.treatsub,aes(x= MR15.cor, y=lifespan, colour=Host))+geom_point()+stat_smooth(method="loess") + facet_grid(.~treatment)

ggplot(data5.15.treatsub,aes(x=MR15.cor, y=lifespan, colour=Host))+geom_point()+stat_smooth(method="lm") + facet_grid(.~treatment)

ggplot(data5.15.treatsub,aes(x=log10(MR15.cor), y=lifespan, colour=Host))+geom_point()+stat_smooth(method="lm")
```

#Negative binomial regression to determine the effect of metabolic rate on lifespan between hosts and its' treatments
```{r}
#separating out RT and SO 
RT15 <- data5.15.treatsub%>%
    filter(treatment=="RT")

SO15 <- data5.15.treatsub%>%
    filter(treatment=="SO")

#RT summary (no stat significance)
mod3<- glm.nb(lifespan~MR15.cor*Host + mass_day14, data=RT15)
summary(mod3)

mod3.1<-glm.nb(lifespan~msMR15.cor*Host, data=RT15)
summary(mod3.1)

#SO summary : HostHaw (in reference to apple) - as lifespan increases, the MR decreases by the estimate (logvalue)
  #mod4 shows interaction b/w host and mr
mod4<-glm.nb(lifespan~MR15.cor*Host+mass_day14, data=SO15)
summary(mod4)

mod4.1<-glm.nb(lifespan~msMR15.cor*Host, data=SO15)
summary(mod4.1)

#mod6 shows how mr and host individually relate to lifespan
mod6<-glm.nb(lifespan~MR15.cor + Host, data=SO15)
summary(mod6)
```

*Likelihood ratio tests of Negative Binomial Models*
```{r}
m1 <- update(mod4, . ~ . - prog)
anova(mod4, m1)

#Checking model assumption -- is negative binomial regression (NBR) a good model for this data?
##Values close to 0 (estimates the dispersion parameter) strongly suggest the NBR model is more appropriate than the Poisson model
mod5 <- glm(lifespan ~ Host*MR15.cor, family = "poisson", data = SO15)
pchisq(2 * (logLik(mod4) - logLik(mod5)), df = 1, lower.tail = FALSE)

#output below indicates that Hawthorne is 0.432 times less likely to live than apple
#confidence interval
(est <- cbind(Estimate = coef(mod4), confint(mod4)))
#incident rate ratio - lifespan ratio
exp(est)

#Filter out all the NAs in order to have same dimensions for predvalues
filSO15<- SO15%>%
  filter(MR15.cor!= "NA"&lifespan !="NA")

#Calculate the predicted values
filSO15$predvalues <- predict(mod6, type = "response")
glimpse(filSO15)


#overlay of real data(points)  with predicted values(line)
a<-ggplot(filSO15,aes(x=MR15.cor, y=lifespan, colour=Host))+geom_point()+geom_line(data=filSO15, aes(x=MR15.cor, y=predvalues, colour=Host))
```

# Double Checking Eclosion for Pupal Deaths
```{r}
#Assign names to each column and column bind in order to simplify and find data easily 
host <- data[,11]
cd <- data[,10]
wid <- data[,29]
unid<-data[,26]
edate <- data[,27]
dcheck<-cbind(host, cd, wid, unid, edate)
```

#Merging pupal death data sheet lifespans
Data available on [github](https://raw.githubusercontent.com/adnguyen/Circadian_rhythm_runs_seasonal_timing/master/Data/2018-06-04_pupal_deaths_data.csv)
```{r}
pupdeaths<- fread("https://raw.githubusercontent.com/adnguyen/Circadian_rhythm_runs_seasonal_timing/master/Data/2018-06-04_pupal_deaths_data.csv")

#Filter out non-eclosers
data5.na <- data5.15%>%
  mutate(neweclosions=as.numeric(neweclosions))%>%
  #glimpse()
  filter(is.na(neweclosions))

#Filter out individuals with adult lifespans
data5.ls <- data5.15.treatsub%>%
  filter(!is.na(lifespan))

#Merge the two sets of data (lifespan and non-eclosers) by "Host", "cohort_day", "well_id", and "uniqueID"
mergedat1 <- inner_join(pupdeaths, data5.na, by = c("Host", "cohort_day", "well_id", "uniqueID"))

#column bind the merged data with the lifespan data to check same column dim
cbind(names(data5.ls), names(mergedat1)[-56])

#once the column dimensions are the same, merge the two sets of data
mergedata<- merge(data5.ls, mergedat1)

#Finally, row bind the data with the desired conditions
finalmerge<- rbind(data5.ls[,c("Host", "cohort_day", "well_id", "uniqueID", "lifespan", "MR15.cor", "treatment", "mass_day14")], mergedat1[,c("Host", "cohort_day", "well_id", "uniqueID", "lifespan", "MR15.cor", "treatment", "mass_day14")])

glimpse(finalmerge)

```

#Negative binomial regression reanalyzed with pupal death data
```{r}
RT15.n <- finalmerge%>%
    filter(treatment=="RT")

SO15.n <- finalmerge%>%
    filter(treatment=="SO")

#RT summary (no stat significance)
mod3.n<- glm.nb(lifespan~MR15.cor*Host, data=RT15.n)
summary(mod3.n)

#summary : HostHaw (in reference to apple) - as lifespan increases, the MR decreases by the estimate (logvalue)
#Coefficient describes relationship, i.e. for mod4SO, higher mass=higher life span according to this model (neg. binomial regression) (p value high=supports null hypothesis)
  #mod4 shows interaction b/w host and mr
mod4.nRT<-glm.nb(lifespan~MR15.cor*Host + mass_day14, data=RT15.n)
summary(mod4.nRT)

mod4.nSO<-glm.nb(lifespan~MR15.cor*Host + mass_day14, data=SO15.n)
summary(mod4.nSO)

#mod6 shows how mr and host individually relate to lifespan
mod6.nSO<-glm.nb(lifespan~MR15.cor + Host + mass_day14, data=SO15.n)
summary(mod6.nSO)

mod6.nRT<-glm.nb(lifespan~MR15.cor + Host + mass_day14, data=RT15.n)
summary(mod6.nRT)

#Likelihood ratio tests of Negative Binomial Models
m1.nRT <- update(mod4.nRT, . ~ . - prog)
anova(mod4.nRT, m1.n)

#Checking model assumption -- is negative binomial regression (NBR) a good model for this data?
##Values close to 0 (estimates the dispersion parameter) strongly suggest the NBR model is more appropriate than the Poisson model
mod5.n <- glm(lifespan ~ Host*MR15.cor, family = "poisson", data = SO15.n)
pchisq(2 * (logLik(mod4.n) - logLik(mod5.n)), df = 1, lower.tail = FALSE)

#output below indicates that Hawthorne is 0.432 times less likely to live than apple
#confidence interval
(est <- cbind(Estimate = coef(mod4.nRT), confint(mod4.nRT)))
#incident rate ratio - lifespan ratio
exp(est)

#Filter out all the NAs in order to have same dimensions for predvalues
filSO15.n<- SO15.n%>%
  filter(MR15.cor!= "NA"&lifespan !="NA")

filRT15.n<- RT15.n%>%
  filter(MR15.cor!= "NA"&lifespan !="NA")

#Calculate the predicted values
filSO15.n$predvalues <- predict(mod4.nSO, type = "response")
glimpse(filSO15.n)

filRT15.n$predvalues <- predict(mod4.nRT, type = "response")
glimpse(filSO15.n)


#overlay of real data(points)  with predicted values(line)
ggplot(filSO15.n,aes(x=MR15.cor, y=lifespan, colour=Host))+geom_point()+geom_line(data=filSO15.n, aes(x=MR15.cor, y=predvalues, colour=Host))

ggplot(filRT15.n,aes(x=MR15.cor, y=lifespan, colour=Host))+geom_point()+geom_line(data=filRT15.n, aes(x=MR15.cor, y=predvalues, colour=Host))

#comparison of non-pupal deaths w/ pupal deaths
grid.arrange(nrow=2,a,b)
```

# Kaplan-Meier Survival Estimate
```{r}
#install.packages(c("survival", "survminer"))
#install.packages(c("ggpubr", "magrittr"))


#Assign censored data and create new column
finalmerge$status[finalmerge$lifespan >= 0] <- 1
fmerge <- finalmerge %>%
      filter(treatment != "")


#Normalize data
#install.packages("BBmisc")
n.finalmerge<-normalize(finalmerge [], method = "standardize", range = c(0, 1), margin = 1L, on.constant = "quiet")

#compute kaplan-Meier survival estimate-so compute the survival probability by host & treatment:
  #fit1 computes survival probability by Host only
  #fit2 computes survival probability by Treatment only
  #fit3 computes survival probability by Host + Treatment

fit1 <- survfit(Surv(lifespan, status) ~ Host, data = finalmerge)
print(fit1)

fit2 <- survfit(Surv(lifespan, status) ~ treatment, data = finalmerge)
print(fit2)

fit3 <- survfit(Surv(lifespan, status) ~ Host + treatment, data = finalmerge)
print(fit3)

# Summary of survival curves
summary(fit1) 
summary(fit2)
summary(fit3)

# Access to the sort summary table
summary(fit1)$table
summary(fit2)$table
summary(fit3)$table

# Visualize curves (Change color, linetype by strata, risk.table color by strata)
ggsurvplot(fit1,
          pval = TRUE, conf.int = TRUE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#E7B800", "#2E9FDF"),
          xlab = "Lifespan in days",
           legend.title = "Conditions",
          pval.size=4)

ggsurvplot(fit2,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = "strata", # Change line type by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#DBAECB", "#2E9FDF","#E7B800"),
          xlab = "Lifespan in days",
          legend.title = "Conditions",
          pval.size=4)

ggsurvplot(fit3,
          pval = TRUE, conf.int = FALSE,
          risk.table = TRUE, # Add risk table
          risk.table.col = "strata", # Change risk table color by groups
          linetype = c("dashed","solid", "dashed", "dashed","solid"), # Change line type                                                                               by groups
          surv.median.line = "hv", # Specify median survival
          ggtheme = theme_bw(), # Change ggplot2 theme
          palette = c("#F9E996","#FFD700","#A9A9A9","#496DA1","#204A87"),
          xlab = "Lifespan in days",
          legend.title = "Conditions",
          pval.size=4)
```

#Cox Regression
```{r}
#univariate
res.cox1 <- coxph(Surv(lifespan, status) ~ Host, data = fmerge)
res.cox1
summary(res.cox1)

res.cox2 <- coxph(Surv(lifespan, status) ~ treatment, data = fmerge)
res.cox2
summary(res.cox2)

#multivariate
covariates <- c("host", "treatment", "MR15.cor","mass_day14")
#For RT
univ_formulasRT <- sapply(covariates,
                        function(x) as.formula(paste('Surv(lifespan, status)~', fmergeRT)))
#For SO
univ_formulasSO <- sapply(covariates,
                        function(x) as.formula(paste('Surv(lifespan, status)~', fmergeSO)))
                        
univ_modelsRT <- lapply(univ_formulasRT, function(x){coxph(x, data = fmergeRT)})
univ_modelsSO <- lapply(univ_formulasSO, function(x){coxph(x, data = fmergeSO)})

# Extract data 
#RT
univ_resultsRT <- lapply(univ_modelsRT,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=2)
                          wald.test<-signif(x$wald["test"], digits=2)
                          beta<-signif(x$coef[1], digits=2);#coeficient beta
                          HR <-signif(x$coef[2], digits=2);#exp(beta)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                          HR <- paste0(HR, " (", 
                                       HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, wald.test, p.value)
                          names(res)<-c("beta", "HR (95% CI for HR)", "wald.test", 
                                        "p.value")
                          return(res)
                          #return(exp(cbind(coef(x),confint(x))))
                         })
#SO
univ_resultsSO <- lapply(univ_modelsSO,
                       function(x){ 
                          x <- summary(x)
                          p.value<-signif(x$wald["pvalue"], digits=2)
                          wald.test<-signif(x$wald["test"], digits=2)
                          beta<-signif(x$coef[1], digits=2);#coeficient beta
                          HR <-signif(x$coef[2], digits=2);#exp(beta)
                          HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                          HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                          HR <- paste0(HR, " (", 
                                       HR.confint.lower, "-", HR.confint.upper, ")")
                          res<-c(beta, HR, wald.test, p.value)
                          names(res)<-c("beta", "HR (95% CI for HR)", "wald.test", 
                                        "p.value")
                          return(res)
                          #return(exp(cbind(coef(x),confint(x))))
                         })


resRT <- t(as.data.frame(univ_resultsRT, check.names = FALSE))
as.data.frame(resRT)
resSO <- t(as.data.frame(univ_resultsSO, check.names = FALSE))
as.data.frame(resSO)

#Multivariate case
#Testing interaction between Host, Treatment, and MR -- Shows interaction b/w treatment & Host
res.cox <- coxph(Surv(lifespan, status) ~ Host*treatment*MR15.cor + mass_day14, data =  fmerge)
summary(res.cox)

#Split up treatments into: SO & RT
fmergeSO <- fmerge%>%
  filter(treatment == "SO")

fmergeRT <- fmerge%>%
  filter(treatment == "RT")


#Coefficient describes relationship: for Cox Regression, coefficient describes relationship to likelihood of death, so neg coefficient for RTmass, lower masses will be more likely to die
#Test interaction between Host, RT, and MR
res.coxRT <- coxph(Surv(lifespan, status) ~ Host*MR15.cor + mass_day14, data =  fmergeRT)
summary(res.coxRT)

#Test interaction between Host, SO, and MR
res.coxSO <- coxph(Surv(lifespan, status) ~ Host*MR15.cor + mass_day14, data =  fmergeSO)
summary(res.coxSO)



#Visualize the estimated distribution of lifespan
  # Plot the baseline survival function
ggsurvplot(survfit(res.coxRT),
           ggtheme = theme_minimal(),
           xlab = "lifespan")


ggsurvplot(data= finalmerge, survfit(res.cox1), color = "#2E9FDF",
           ggtheme = theme_minimal(),
           xlab = "lifespan")

ggsurvplot(data= finalmerge, survfit(res.cox2), color = "#2E9FDF",
           ggtheme = theme_minimal(),
           xlab = "lifespan")

#New data, with the impact of treatment only
treatment_df <- with(finalmerge,
               data.frame(treatment = c(1, 1), 
                          lifespan = c(1, 1),
                          Host = c(1,2)
                          )
               )
treatment_df

# Survival curves
fit <- survfit(res.cox, newdata = treatment_df)
ggsurvplot(fit, conf.int = TRUE, legend.labs=c("Host=Apple", "Host=Haw"),
           ggtheme = theme_minimal())

```

# Session Info
```{r}
sessionInfo()
```