---
title: "viability_selection_mr_lifespan"
output: 
  html_document: 
    toc: yes
---

#Libraries
```{r}
library(dplyr) #data parsing library
library(data.table) #data parsing library mainly for reading and writing out
library(ggplot2) #data visualization library
library(lubridate) #package to deal with time variables
library(curl)
```

# Download data and do preliminary parsing
Data are publicly available via [github](https://raw.githubusercontent.com/adnguyen/Circadian_rhythm_runs_seasonal_timing/master/Data/2018-05-30_rhagoletis_masterdata_data_slice.csv)
```{r}
data<- fread("https://raw.githubusercontent.com/adnguyen/Circadian_rhythm_runs_seasonal_timing/master/Data/2018-05-30_rhagoletis_masterdata_data_slice.csv")

#look at summary of data
glimpse(data)
```

#Calculating metabolic rates
```{r}
glimpse(data$purge_time_1)
hm(data$purge_time_1)
data$day10purge <- lubridate::hour(hm(data$purge_time_1))+lubridate::minute(hm(data$purge_time_1))/60

#figure out how to create a time sequence
start<-13.26667
end<- 14.5
n=75
seq(start,end, length.out = n)

#Getting start and end (min and max) of purges and smaple size for each host, cohort day, and tape
param <- data%>%
  group_by(cohort_day, tape)%>%
  summarise(max=max(day10purge, na.rm=TRUE), min=min(day10purge, na.rm=TRUE), n=length(cohort_day))

#goal: for this section, we want a sequence of times for day 10 purge
data2 <- data%>%
  group_by(cohort_day, tape)%>%
  mutate(.,day10purge.trans=seq(from = min(day10purge, na.rm=TRUE), to = max(day10purge, na.rm=TRUE), length.out = length(Host)))
glimpse(data2)

#cohort 2&3 for apple have the right sequence of purge times so they(day10purge) need to be replaced into corresponding day10purge.trans column
#extracting number of rows we want to replace
data2[1:500,]%>%
  filter(cohort_day < 4)%>%
  dim()
data2[1:211,46]<- data2[1:211,45]
data2[1:211,46]

#calculating start and end time
glimpse(data$resp_time_1)
hms(data$resp_time_1)
data2$day10resp <- lubridate::hour(hms(data$resp_time_1))+lubridate::minute(hms(data$resp_time_1))/60

data2$total_time_day10 <- (24 - data2$day10purge.trans) + data2$day10resp

#getting denominator for mass-specific mr
data2$total_time_day10 * data$mass_day10

data2$MR11<- data$resp_day11/(data2$total_time_day10)
data2$msMR11<- data$resp_day11/(data2$total_time_day10 * data$mass_day10)

#need to control for blank controls
data3 <- data2%>%
  group_by(cohort_day, tape)%>%
  filter(Site_name=="Blank")%>%
  summarise(mean.blank=mean(MR11,na.rm=TRUE))
#check columns
glimpse(data3)
data3$mean.blank

#merge data3 and data2 by cohort day and tape
data4 <- inner_join(data2, data3, by=c("cohort_day", "tape"))
data4
data4$mean.blank

#do some corrections
data5 <- data4%>%
  mutate(MR11.cor = MR11 - mean.blank, msMR11.cor = msMR11 - mean.blank)
glimpse(data5)

data5.neg <- data5%>% 
  filter(MR11.cor<0)%>%
  dim()
  #select(.,c("cohort_day","tape", "uniqueID", "MR11.cor"))
   #fwrite(data5.neg,"2018-05-31_negative_MR.csv")

#ggplot(data5.neg, aes(x=MR11.cor))+geom_histogram()
```

## Figures of mr between hosts
```{r}
ggplot(data5, aes(x=Host, y=MR11.cor))+geom_boxplot()
ggplot(data5, aes(x=MR11.cor, fill=(Host)))+geom_histogram(position = "identity", alpha=.5)
```
#Calculate eclosion days
```{r}
data5$neweclosions<-difftime(as.Date(data5$eclosion_date), as.Date(data5$Eclosion_reference_date),units=c("days"))+15
```
#Explore eclosion data
```{r}
ggplot(data5.treatsub, aes(x=MR11.cor, fill=(Host)))+geom_histogram(position = "identity", alpha=.5)+facet_grid(.~treatment)

a<-ggplot(data5.treatsub, aes(x=MR11.cor, fill=(Host)))+geom_density(position = "identity", alpha=.5)+facet_grid(.~treatment)

b<-ggplot(data5.treatsub, aes(x=neweclosions, fill=(Host)))+geom_density(position = "identity", alpha=.5)+facet_grid(.~treatment)

#4 panel
install.packages("gridExtra")
library(gridExtra)
grid.arrange(nrow=2,a,b)

```
###Test differences in eclosion timing between host and experiment
```{r}
mod1<- aov(neweclosions~Host*treatment, data=data5.treatsub)
summary(mod1)

#narrow down to SO
data5.SO<-data5.treatsub%>%
  filter(treatment=="SO")

mod2<-aov(neweclosions~Host, data=data5.SO)
summary(mod2)

#how balanced it is
data5.SO%>%
  group_by(Host)%>%
  summarise(n=length(Host))
```
#Explore relationship between eclosion days and MR
**scatter plots**
```{r}

data5.treatsub<-data5%>%
  filter(treatment!="GC"&treatment!="")

ggplot(data5.treatsub,aes(x=MR11.cor, y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="loess")+facet_grid(.~treatment)

#plotting mr11 on the x axis and new eclosions on the y axis; color points by host, fit curve to data
ggplot(data5.treatsub,aes(x=MR11.cor, y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="loess")


#same - fit linear curve to data
ggplot(data5.treatsub,aes(x=MR11.cor, y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="lm")

#could log transform mr to make it more linear
ggplot(data5.treatsub,aes(x=log10(MR11.cor), y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="lm") + xlim(-3.2,-.5)
```

#Calculating lifespan
```{r}
data5.treatsub$lifespan<-difftime(as.Date(data5.treatsub$Adult_death_date, na.rm=TRUE), as.Date(data5.treatsub$eclosion_date, na.rm=TRUE),units=c("days"))

#finding negative lifespans
lsneg <- data5.treatsub%>% 
  filter(lifespan < 0) %>%
  select(uniqueID,cohort_day, tape)

```

## Figures of lifespan between hosts
```{r}

ggplot(data5.treatsub, aes(x=Host, y=lifespan))+geom_boxplot()
ggplot(data5.treatsub, aes(x=lifespan, fill=(Host)))+geom_histogram(position = "identity", alpha=.5)

```

#Associating MR with lifespan
```{r}

ggplot(data5.treatsub,aes(x= MR11.cor, y=lifespan, colour=Host))+geom_point()+stat_smooth(method="loess") + facet_grid(.~treatment)

a<- ggplot(data5.treatsub,aes(x=MR11.cor, y=lifespan, colour=Host))+geom_point()+stat_smooth(method="lm") + facet_grid(.~treatment)

ggplot(data5.treatsub,aes(x=log10(MR11.cor), y=lifespan, colour=Host))+geom_point()+stat_smooth(method="loess") + xlim(-3.2,-.5) + ylim(-.5, 150)

```


## Repeat above, but with day 15
```{r}
glimpse(data$purge_time_2)
hm(data$purge_time_2)
data$day15purge <- lubridate::hour(hm(data$purge_time_2))+lubridate::minute(hm(data$purge_time_2))/60

#figure out how to create a time sequence
start<-13.26667
end<- 14.5
n=75
seq(start,end, length.out = n)

#Getting start and end (min and max) of purges and sample size for each host, cohort day, and tape
param <- data%>%
  group_by(cohort_day, tape)%>%
  summarise(max=max(day15purge, na.rm=TRUE), min=min(day10purge, na.rm=TRUE), n=length(cohort_day))

#goal: for this section, we want a sequence of times for day 10 purge
data2.15 <- data%>%
  group_by(cohort_day, tape)%>%
  mutate(.,day15purge.trans=seq(from = min(day15purge, na.rm=TRUE), to = max(day15purge, na.rm=TRUE), length.out = length(Host)))
glimpse(data2.15)

#calculating start and end time
glimpse(data$resp_time_2)
hms(data$resp_time_2)
data2.15$day15resp <- lubridate::hour(hms(data$resp_time_2))+lubridate::minute(hms(data$resp_time_2))/60

data2.15$total_time_day15 <- (24 - data2.15$day15purge.trans) + data2.15$day15resp

#getting denominator for mass-specific mr
data2.15$total_time_day15 * data$mass_day14

data2.15$MR15<- data$resp_day15/(data2.15$total_time_day15)
data2.15$msMR15<- data$resp_day15/(data2.15$total_time_day15 * data$mass_day14)

#need to control for blank controls
data3.15 <- data2.15%>%
  group_by(cohort_day, tape)%>%
  filter(Site_name=="Blank")%>%
  summarise(mean.blank2=mean(MR15,na.rm=TRUE))
#check columns
glimpse(data3.15)
data3.15$mean.blank2

#merge data3 and data2 by cohort day and tape
data4.15 <- inner_join(data2.15, data3.15, by=c("cohort_day", "tape"))
data4.15
data4.15$mean.blank2

#do some corrections
data5.15 <- data4.15%>%
  mutate(MR15.cor = MR15 - mean.blank2, msMR15.cor = msMR15 - mean.blank2)
glimpse(data5.15)

data5.15.neg <- data5.15%>% 
  filter(MR15.cor<0)%>%

  #select(.,c("cohort_day","tape", "uniqueID", "MR11.cor"))
   #fwrite(data5.neg,"2018-05-31_negative_MR.csv")

#ggplot(data5.neg, aes(x=MR11.cor))+geom_histogram()


#figures of mr b/w hosts

ggplot(data5.15, aes(x=Host, y=MR15.cor))+geom_boxplot()

ggplot(data5.15, aes(x=MR15.cor, fill=(Host)))+geom_histogram(position = "identity", alpha=.5)
#eclosion data
data5.15.treatsub$neweclosions<-difftime(as.Date(data5.15.treatsub$eclosion_date), as.Date(data5.15.treatsub$Eclosion_reference_date),units=c("days"))+15

#scatter plots
data5.15.treatsub<-data5.15%>%
  filter(treatment!="GC"&treatment!="")

ggplot(data5.15.treatsub, aes(x=MR15.cor, y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="loess")+facet_grid(.~treatment)

#plotting mr15 on the x axis and new eclosions on the y axis; color points by host, fit curve to data
ggplot(data5.15.treatsub,aes(x=MR15.cor, y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="loess")

#same - fit linear curve to data
ggplot(data5.15.treatsub,aes(x=MR15.cor, y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="lm")

#could log transform mr to make it more linear
ggplot(data5.15.treatsub,aes(x=log10(MR15.cor), y=neweclosions, colour=Host))+geom_point()+stat_smooth(method="lm") + xlim(-3.2,-.5)


#mr15 w lifespan
data5.15.treatsub$lifespan<-difftime(as.Date(data5.15.treatsub$Adult_death_date, na.rm=TRUE), as.Date(data5.15.treatsub$eclosion_date, na.rm=TRUE),units=c("days"))

ggplot(data5.15.treatsub,aes(x= MR15.cor, y=lifespan, colour=Host))+geom_point()+stat_smooth(method="loess") + facet_grid(.~treatment)

ggplot(data5.15.treatsub,aes(x=MR15.cor, y=lifespan, colour=Host))+geom_point()+stat_smooth(method="lm") + facet_grid(.~treatment)

ggplot(data5.15.treatsub,aes(x=log10(MR15.cor), y=lifespan, colour=Host))+geom_point()+stat_smooth(method="lm")

grid.arrange(nrow=2,a,b)
```

#negative binomial regression to determine the effect of metabolic rate on lifespan between hosts and its' treatments
```{r}
library(MASS)
#determining class
class(data5.15.treatsub$lifespan)
data5.15.treatsub$lifespan<-as.numeric(data5.15.treatsub$lifespan)

#model2<-glm.nb(lifespan~MR15.cor*Host*treatment, data=data5.15.treatsub)


#separating out RT and SO 
RT15 <- data5.15.treatsub%>%
    filter(treatment=="RT")

SO15 <- data5.15.treatsub%>%
    filter(treatment=="SO")

#RT summary (no stat significance)
mod3<- glm.nb(lifespan~MR15.cor*Host, data=RT15)
summary(mod3)

#SO summary : HostHaw (in reference to apple) - as lifespan increases, the MR decreases by the estimate (logvalue)
mod4<-glm.nb(lifespan~MR15.cor*Host, data=SO15)
summary(mod4)

```

#Session Info
```{r}
sessionInfo()
```